Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2020-01-20 aosd13/paperBody.tex
Fix typo
WIP increase code size, more overfull hboxes in figures
Tune figure setup
Code line breaks: Replace \linebreak with better \codebreak{,Space}

A call to \code{optimize} on a query will then take this index into account and rewrite the query to perform index lookup instead of scanning, if possible. For instance, the code in Fig.~\ref{fig:reifiedQuery} would be transparently rewritten by the optimizer to a query similar to the following:
\begin{lstlisting}
val indexedQuery =
  for {
    book <- idxByPublisher("Pearson Education")
    author <- book.authors
  } yield new BookData(book.title,
    author.firstName + " " + author.lastName,
    book.authors.size - 1)
\end{lstlisting}
Since dictionaries in Scala are functions, in the above code, dictionary lookup on \code{idxByPublisher} is represented simply as function application. The above code iterates over books having the desired publisher, instead of scanning the whole library, and performs the remaining computation from the original query. Although the index use in the listing above is written as
%Beware: \linebreak simulates a normal break, hence it doesn't cause an
%underfull hbox. See https://tex.stackexchange.com/questions/39629/why-is-linebreak-failing#comment80030_39629.
\code{idxByPublisher("Pearson}\linebreak\code{Education")}, only the cached result of evaluating the index is used when
the query is executed, not the reified index definition.

@@

More in detail, our compilation algorithm is as follows.
(a) We implement CSP by replacing embedded Scala values by references to the
function arguments; so for instance \code{List(1, 2, 3).map(}\linebreak\code{x => x + 1)} becomes
the function \code{(s1: List[Int], s2: Int) => s1.map(x => x + s2)}.
(b) We look up the produced expression tree, together with the types of the
constants we just removed, in a cache mapping to the generated classes. If the
lookup fails we update the cache with the result of the next steps.
(c) We apply CSE on the expression.
(d) We convert the tree to code, compile it and load the generated code.

>>>>

A call to \code{optimize} on a query will then take this index into account and rewrite the query to perform index lookup instead of scanning, if possible. For instance, the code in Fig.~\ref{fig:reifiedQuery} would be transparently rewritten by the optimizer to a query similar to the following:
\begin{lstlisting}
val indexedQuery =
  for {
    book <- idxByPublisher("Pearson Education")
    author <- book.authors
  } yield new BookData(book.title,
    author.firstName + " " + author.lastName,
    book.authors.size - 1)
\end{lstlisting}
Since dictionaries in Scala are functions, in the above code, dictionary lookup on \code{idxByPublisher} is represented simply as function application. The above code iterates over books having the desired publisher, instead of scanning the whole library, and performs the remaining computation from the original query. Although the index use in the listing above is written as
\code{idxByPublisher(}\codebreak\code{"Pearson Education")}, only the cached result of evaluating the index is used when
the query is executed, not the reified index definition.

@@

More in detail, our compilation algorithm is as follows.
(a) We implement CSP by replacing embedded Scala values by references to the
function arguments; so for instance \code{List(1, 2, 3).map(}\codebreak\code{x => x + 1)} becomes
the function \code{(s1: List[Int], s2: Int) => s1.map(x => x + s2)}.
(b) We look up the produced expression tree, together with the types of the
constants we just removed, in a cache mapping to the generated classes. If the
lookup fails we update the cache with the result of the next steps.
(c) We apply CSE on the expression.
(d) We convert the tree to code, compile it and load the generated code.

### 2020-01-20 aosd13/pepm13/caseStudyImpl.tex
Code line breaks: Replace \linebreak with better \codebreak{,Space}

With these declarations in scope, the Scala compiler rewrites \code{str1 + str2} to
\code{expToStringOps(}\linebreak\code{str1).+(str2)}, which evaluates to \code{StringConcat(str1, str2)} as desired.
The implicit conversion function \code{expToStringOps} is not applicable to \code{Exp[Boolean]} because it explicitly specifies the receiver of the \code{+}-call to have type \code{Exp[String]}.
In other words, expressions like \code{str1 + str2} are now \emph{lifted} on the level of expression trees in a type-safe way. For brevity, we refer to the defined operator as \code{Exp[String].+}.

\smartParagraph{Literal values}
However, a string concatenation might also include constants, as in \code{str1 + "foo"} or \code{"bar" + str1}.
To lift \code{str1 + "foo"}, we introduce a lifting for constants which wraps them in a \code{Const} node:

@@

Implicit conversions for user-defined types cooperate with other ones; for instance, \code{author.}\linebreak\code{firstName + " " + author.lastName} is rewritten to
\begin{lstlisting}
(expToStringOps(expToAuthorOps(author).firstName) + pure(" ")) +
  expToAuthorOps(author).lastName
\end{lstlisting}

@@

%\pg{Select and integrate what needed.}
\smartParagraph{Manipulating functions}
%\section*{More about functions.}
%\tr{This material is left over from Sec. 4.3. It is partly needed in later sections and needs to be moved there.}
To be able to inspect a HOAS function body \code{funBody: Exp[S] =>}\linebreak\code{Exp[T]}, like \code{str => str + "!"}, we convert it to \emph{first-order} abstract syntax (FOAS), that is to an expression tree of type \code{Exp[T]}.
To this end, we introduce a representation of variables and a generator of fresh ones; since variable names are auto-generated, they are internally represented simply as integers instead of strings for efficiency.

>>>>

With these declarations in scope, the Scala compiler rewrites \code{str1 + str2} to
\code{expToStringOps(}\codebreak\code{str1).+(str2)}, which evaluates to \code{StringConcat(str1, str2)} as desired.
The implicit conversion function \code{expToStringOps} is not applicable to \code{Exp[Boolean]} because it explicitly specifies the receiver of the \code{+}-call to have type \code{Exp[String]}.
In other words, expressions like \code{str1 + str2} are now \emph{lifted} on the level of expression trees in a type-safe way. For brevity, we refer to the defined operator as \code{Exp[String].+}.

\smartParagraph{Literal values}
However, string concatenation might also be applied to constants, as in \code{str1 + "foo"} or \code{"bar" + str1}.
To lift \code{str1 + "foo"}, we introduce a lifting for constants which wraps them in a \code{Const} node:

@@

Implicit conversions for user-defined types cooperate with other ones; for instance, \code{author.}\codebreak\code{firstName + " " + author.lastName} is rewritten to
\begin{lstlisting}
(expToStringOps(expToAuthorOps(author).firstName) + pure(" ")) +
  expToAuthorOps(author).lastName
\end{lstlisting}

@@

%\pg{Select and integrate what needed.}
\smartParagraph{Manipulating functions}
%\section*{More about functions.}
%\tr{This material is left over from Sec. 4.3. It is partly needed in later sections and needs to be moved there.}
To be able to inspect a HOAS function body \code{funBody: Exp[S] =>}\codebreakSpace\code{Exp[T]}, like \code{str => str + "!"}, we convert it to \emph{first-order} abstract syntax (FOAS), that is to an expression tree of type \code{Exp[T]}.
To this end, we introduce a representation of variables and a generator of fresh ones; since variable names are auto-generated, they are internally represented simply as integers instead of strings for efficiency.
